
\chapter[Interface to MiniZinc]{{\tt minizinc}: The XSB Interface to MiniZinc-based Constraint
  Solving}

\begin{center}
  {\Large {\bf By Michael Kifer}}
\end{center}



\section{Introduction}

MiniZinc is a uniform declarative constraint language that is understood by
most modern solvers for constraint and optimization problems. It comes bundled with a few such solvers, one of
which, \texttt{gecode}, is top-notch: very powerful and fast.  Other solvers,
including most of the newest ones, can be downloaded separately and
installed as plugins.

The MiniZinc language is described in
\url{https://www.minizinc.org/doc-2.2.3/en/index.html}; see, especially, the
tutorial. 
The XSB interface to MiniZinc comes with several sample problems from that
tutorial, which are found in \texttt{.../XSB/packages/minizinc/examples/}.
The file \texttt{.../XSB/packages/minizinc/examples.P} contains examples of
XSB invocations of those problems. After the installation, all examples can be run
by simply starting XSB and then
%% 
\begin{verbatim}
   | ?- [minizinc].   %% load the package
   | ?- [examples].   %% run all examples
\end{verbatim}
%% 

\section{Installation}

Some Linux distributions (e.g., Ubuntu) come with ready-made MiniZinc .deb
or .rpm packages. However, one must make sure that the command
\url{minizinc} is provided by those packages (some provide only the IDE).
Mac packages are also available.

In case a Linux or a Mac package is incomplete (or if one uses Windows),
MiniZinc can be downloaded from
\url{https://www.minizinc.org/software.html}. 

Once installed, make sure that the command \texttt{minizinc} is understood when typed in a
command window. If not, add \emph{folder-to-where-}\textbf{/bin/}minizinc is sitting to the
environment variable \texttt{PATH}.  In Windows, this is best done in
Control Panel; in Linux and Mac, add the command
%% 
\begin{verbatim}
    export PATH=$PATH:path-to-minizinc
\end{verbatim}
%% 
to \texttt{.bashrc} or an equivalent place. For instance,
%% 
\begin{verbatim}
    PATH=$PATH:$HOME/minizinc/MiniZincIDE-2.2.3-bundle-linux/bin
\end{verbatim}
%% 

\section{The API}

Constraint and optimization problems are specified using the MiniZinc
language in \emph{model files}, which have the suffix \texttt{.mzn}.
Such problems usually have a number of \emph{input variables}   and several
\emph{output} (or \emph{decision}) \emph{variables}.   
In principle, input values can be specified in the model file itself, but
this is generally is not a good idea because typically one wants to solve
the same problem with different outputs.
For this reason, MiniZinc allows one to invoke a model (which has no input)
on one or more \emph{input files}, which have the extension \texttt{.dzn}. 
In addition, XSB's API to MiniZinc lets one pass parameters in-line, as
part of a Prolog call.

The API itself mainly consists of the following calls, which live in XSB's
module \texttt{minizinc}: 
%% 
\begin{itemize}
\item
  \texttt{solve(+MznF,+DatFs,+InPars,+Solver,+Solns,+OutTempl,-Rslt,-Xceptns)}: 
  The meaning of the arguments is as follows:
  %% 
  \begin{itemize}
  \item    \texttt{MznF}: should be bound to a path to to the desired MznF
    model file (.mzn file) that contains a specification of a constraint or
    optimization problem. The path must be represented as a Prolog atom and
    can be absolute or relative to the current directory.
  \item \texttt{DatFs}: a list of paths (relative or absolute)  
    to the data files that describe all or some of the input parameters for
    the model in \texttt{MznF}.
    If no data files are needed, just use the empty list \texttt{[]}. 
  \item \texttt{InPars}: a list of the \emph{in-line} input parameters.
    These are supported for flexibility, to allow initialization of some or
    all input parameters directly in Prolog.
    Each parameter in \texttt{InPars} must have the form \emph{id} =
    \emph{value}, where \emph{id} must be the Id of an input variable used in
    the model \texttt{MznF} and value must be a term understood by MiniZinc.
    Typically, such a term would be a number, an atom, or a function term.
    For instance,  \texttt{foo='[|1 0|3 4|8 9|]'} would initialize the input
    variable \texttt{foo} with a 2-dimensional array of numbers;
    \texttt{'Item' = anon\_enum(8)}  would initialize the MiniZinc variable
    \texttt{Item}  of an enumerated type to a set
    \texttt{\{Item\_1,Item\_2,...,Item\_8\}}.  Note that \texttt{Item} must
    be quoted on the Prolog side, since otherwise it would be interpreted
    as a variable.
  \item \texttt{Solver}: the name of the solver to use. If this is a variable,
    the default solver \texttt{gecode} is used. 
  \item \texttt{Solns}: the number of solutions to show. 
    Could be \texttt{all} or a positive integer.
    Note: for optimization problems where a function is maximized or minimized,
    \texttt{Solns} \underline{must} be bound to 1. Otherwise, solvers would
    return also non-optimal solutions, and the desired optimal one may not
    be the first one.
  \item \texttt{OutTempl}: the template showing how the output should look
    like. It has the form \emph{predname}($OutSpec_1$, ...,$OutSpec_n$)
    where \emph{predname} is the name of a predicate in \texttt{usermod}
    where the results will be stored  and each $OutSpec$ can have one of
    these forms
    %% 
    \begin{itemize}
    \item  An \emph{atom} representing an output (``decision'') variable
      defined in the model file \texttt{MznF}.
      In the result, this atom will be replaced with the value of that variable.
    \item  A term of the form \texttt{+}(\emph{atom}) or
      \texttt{str}(\emph{atom}).  In the result, this will be replaced with
      \emph{atom} verbatim. Note: if \texttt{atom} is not alphanumerical or
      if it starts with a capital letter, it must be quoted.

    \item A list of the form [$OutSpec_1$, ...,$OutSpec_n$]. Each
      \emph{Spec} in the list will be replaced according to the rules
      described here.
    \item A term of the form $OutSpec_1$ \texttt{=} $OutSpec_2$. Each
      \emph{Spec} will be replaced according to the rules described here.
    \end{itemize}
    %% 
  \item \texttt{Rslt}: must be a term that matches the output
    template.  Solutions to the constraint and optimization problems
    will be returned as bindings to variables in this term. They will
    also be asserted as facts in \texttt{usermod}.
  \item \texttt{Xceptns}: a list of exceptions returned by the solver.
    If the constraint/optimization problem was solved successfully, this
    variable will be bound to an empty list \texttt{[]}. Otherwise,
    each exception has the form
    \texttt{(reason=...,model\_file=...)}  and the
    following reasons might be returned:
    %% 
    \begin{itemize}
    \item  \texttt{unsatisfiable} --- the optimization problem is
      unsatisfiable.
    \item \texttt{unbounded}  --- the optimization problem has an unbounded
      objective function. For example, if the problem is to maximize the
      function and the function has no maximum (under the given
      constraints); or the problem is to minimize the function, and there
      is no minimum.
    \item \texttt{unsatisfiable\_or\_unbounded} --- one of the above.
    \item \texttt{unknown} --- could not find a solution within the limits
      (e.g., timeout).
    \item \texttt{error} --- search resulted in an error. 
    \end{itemize}
    %% 
  \end{itemize}
  %% 
  Note: exceptions are separate from other kinds of errors, such as a
  syntax error in a model or data file or using a solver with a feature or
  option it does not support. Errors are explained later.
\item
  \texttt{solve\_flex(+MznF,+DatFs,+InPars,+Solver,+Solns,+OutTempl,-Rslt,-Xceptns)}: 
  The meaning of the parameters is the same as for \texttt{solve/8}.
  The difference is that if \texttt{InPars} is non-ground, the call to
  MiniZinc is delayed until \texttt{InPars} becomes ground. If the top
  query finishes and \texttt{InPars} is still not ground, MiniZinc will
  \emph{not} be called at all.
  This is used in cases when it is hard to estimate when \texttt{InPars}
  may become ground, so calls to \texttt{solve\_flex} can be placed
  early. But, of course, one must be sure that \texttt{InPars} will get
  bound at some point.  
\end{itemize}
%% 

\paragraph{How does MiniZinc return complex structures back to Prolog?}

\paragraph{Errors vs. exceptions.}
If a model or data file contains a syntax error or a feature that the
chosen solver does not support, the \texttt{solve/8} and
\texttt{solve\_flex/8} calls will fail and a message will be printed to
standard output:
%% 
\begin{verbatim}
    +++ xsb2mzn: syntax or type errors found; details in ....some file...
\end{verbatim}
%% 
The user will then be able to find the details about the problem.

Note: errors are different from exceptions. If only exceptions are returned
(and no errors), the calls \texttt{solve/8} and \texttt{solve\_flex/8} will
succeed. In contrast, they will fail in case of errors.

\paragraph{Debugging API.}
For further development and bug reporting, the following calls are useful.
They are all 0-ary predicates that take no arguments; they all reside in the
XSB module \texttt{minizinc}. 
%% 
\begin{itemize}
\item  \texttt{keep\_tmpfiles}: The MiniZinc interface creates a number of
  temporary files, which are deleted, if MiniZinc finished normally and
  without an error. However, if a bug is suspected, it is desirable to
  preserve these files and send them to the developers. 
  This can be achieved by executing the \texttt{keep\_tmpfiles} predicate
  as a query, before the call to solve.
\item \texttt{show\_mzn\_cmd}: Executing this as a query will cause the
  \texttt{solve/8} predicate, described earlier, to print the shell command
  that is used to invoke MiniZinc for each call.
  This is useful if one suspects a bug in the API.
\item \texttt{dbg\_clear}: Executing this clears out the flags set by the
  above debugging calls. As a result, the temporary files will again be
  deleted after each invocation of MiniZinc and shell commands will not be
  shown.
\end{itemize}
%% 



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "manual2"
%%% End: 
